Accuracy in Parameter Estimation and Simulation Approaches for Sample Size Planning with Multiple Stimuli

### 2023-12-27 manuscript_draft_edits.Rmd
adding this version for edits

ggplot(summary_long %>% 
         group_by(original_n, power, scale_size) %>% 
         summarize(sample_size_avg = mean(sample_size)),
       aes(original_n, sample_size_avg, color = scale_size)) + 
  geom_point() + 
  geom_point(aes(original_n, original_n), color = "black") + 
  geom_line() + 
  theme_linedraw(base_size = 15) + 
  xlab("Pilot Sample Size") + 
  ylab("Projected Sample Size") + 
  scale_color_discrete(name = "Scale Size") +
  facet_wrap(~power)

@@

```{r scale-skew-figure, echo = F, fig.cap="Simulated pilot sample size compared to projected sample size for ceiling versus normal distributions on each scale. " }
ggplot(summary_long %>% 
         group_by(original_n, power, data_type) %>% 
         summarize(sample_size_avg = mean(sample_size)),
       aes(original_n, sample_size_avg, color = data_type)) + 
  geom_point() + 
  geom_point(aes(original_n, original_n), color = "black") + 
  geom_line() + 
  theme_linedraw(base_size = 15) + 
  xlab("Pilot Sample Size") + 
  ylab("Projected Sample Size") + 
  scale_color_discrete(name = "Scale Skew") +
  facet_wrap(~power)
```

@@

```{r scale-hetero-figure, echo = F, fig.cap="Simulated pilot sample size compared to projected sample size for differing amounts of heterogeneity on each scale. "}
ggplot(summary_long %>% 
         group_by(original_n, power, variability) %>% 
         summarize(sample_size_avg = mean(sample_size)),
       aes(original_n, sample_size_avg, color = variability)) + 
  geom_point() + 
  geom_point(aes(original_n, original_n), color = "black") + 
  geom_line() + 
  theme_linedraw() + 
  xlab("Pilot Sample Size") + 
  ylab("Projected Sample Size") + 
  scale_color_discrete(name = "Item Heterogeneity") +
  facet_wrap(~power)
```

@@

```{r corrected-figure, echo = F, fig.cap="Corrected projected sample sizes for variability and power levels. "}
decay <- 1-sqrt((summary_long$original_n-20)/summary_long$original_n)^log2(summary_long$original_n)
summary_long$new_sample <- summary_long$sample_size*decay
ggplot(summary_long %>% 
         group_by(original_n, power, variability) %>% 
         summarize(sample_size_avg = mean(sample_size),
                   new_sample_size_avg = mean(new_sample)),
       aes(original_n, new_sample_size_avg, color = variability)) + 
  geom_point() + 
  geom_point(aes(original_n, original_n), color = "black") + 
  geom_line() + 
  theme_linedraw() + 
  xlab("Pilot Sample Size") + 
  ylab("Projected Sample Size") + 
  scale_color_discrete(name = "Item Heterogeneity") +
  facet_wrap(~power)
```

@@

Last, we examined the question of an appropriate SE decile. First, the 0%, 10%, and 20% deciles are likely too restrictive, providing very large estimates that do not always find a reasonable sample size in proportion to the pilot sample size, scale size, and heterogeneity. If we examine the $R^2$ values for each decile of our regression equation separately, we find that the values are all $R^2$ \> .99 with very little differences between them. Figure \@ref(fig:decile-figure) illustrates the corrected scores for simulations at the the 40% and 50% decile recommended cutoff for item standard errors. For small heterogeneity, differences in decile are minimal, while larger heterogeneity shows more correction at the 40% decile range, especially for scales with larger potential variance. Therefore, we would suggest the 40% decile to overpower each item to determine the minimum and maximum sample size.

```{r decile-figure, fig.cap = "Comparison of the cutoffs for 40% and 50% deciles across heterogeneity (rows), powering of items (columns), and scale size (shape)."}
summary_long %>% 
  filter(name == "Decile 40" | name == "Decile 50") %>% 
  group_by(variability, original_n, scale_size, name, power) %>% 
  summarize(new_sample = mean(new_sample)) %>% 
  ggplot(., aes(original_n, new_sample, color = name)) + 
  geom_point(aes(shape = scale_size)) + 
  geom_point(aes(original_n, original_n), color = "black") + 
  theme_classic(base_size = 10) + 
  facet_wrap(~ variability*power) + 
  xlab("Pilot Sample Size") + 
  ylab("Corrected Sample Size") + 
  scale_shape_discrete(name = "Scale Size") +
  scale_color_discrete(name = "Cutoff Decile") + 
  theme(legend.position="bottom")
```

@@

```{r sim_pop, include = T, echo = T}
# devtools::install_github("SemanticPriming/semanticprimeR")
library(semanticprimeR)
pops <- simulate_population(mu = 4, # item means
  mu_sigma = .2, # variability in item means 
  sigma = 2, # item standard deviations
  sigma_sigma = .2, # standard deviation of the standard deviations
  number_items = 30, # number of items
  number_scores = 20, # number of participants
  smallest_sigma = .02, #* smallest possible standard deviation
  min_score = 1, #* minimum score for truncating purposes
  max_score = 7, #* maximum score for truncating purposes
  digits = 0) #* number of digits for rounding

>>>>

# New facet label names for power variable
power.labs <- c("80% of items", "85% of items",
                "90% of items", "95% of items")
names(power.labs) <- c("80", "85", "90", "95")

plot_colors <- c("#4FEAAC", "#FF826F", "#4F84FD")

ggplot(summary_long %>% 
         group_by(original_n, power, scale_size) %>% 
         summarize(sample_size_avg = mean(sample_size)),
       aes(original_n, sample_size_avg, color = scale_size)) + 
  geom_point() + 
  geom_point(aes(original_n, original_n), color = "black") + 
  geom_line() + 
  geom_line(aes(original_n, original_n), color = "black") + 
  theme_linedraw(base_size = 15) + 
  xlab("\nPilot Sample Size") + 
  ylab("Projected Sample Size\n") + 
  scale_color_manual(name = "Scale Size",
                     values = c(plot_colors)) +
  facet_wrap(~power,
             labeller = labeller(power = power.labs))+
  theme_bw()

@@

```{r scale-skew-figure, echo = F, fig.cap="Simulated pilot sample size compared to projected sample size for ceiling versus normal distributions on each scale. " }
ggplot(summary_long %>% 
         group_by(original_n, power, data_type) %>% 
         summarize(sample_size_avg = mean(sample_size)),
       aes(original_n, sample_size_avg, color = data_type)) + 
  geom_point() + 
  geom_point(aes(original_n, original_n), color = "black") + 
  geom_line() + 
  geom_line(aes(original_n, original_n), color = "black") + 
  theme_linedraw(base_size = 15) + 
  xlab("\nPilot Sample Size") + 
  ylab("Projected Sample Size\n") + 
  scale_color_manual(name = "Scale Skew",
                     values = c(plot_colors)) +
  facet_wrap(~power,
             labeller = labeller(power = power.labs))+
  theme_bw()

@@

```{r scale-hetero-figure, echo = F, fig.cap="Simulated pilot sample size compared to projected sample size for differing amounts of heterogeneity on each scale. "}
ggplot(summary_long %>% 
         group_by(original_n, power, variability) %>% 
         summarize(sample_size_avg = mean(sample_size)),
       aes(original_n, sample_size_avg, color = variability)) + 
  geom_point() + 
  geom_point(aes(original_n, original_n), color = "black") + 
  geom_line() + 
  geom_line(aes(original_n, original_n), color = "black") +
  theme_linedraw() + 
  xlab("\nPilot Sample Size") + 
  ylab("Projected Sample Size\n") + 
  scale_color_manual(name = "Item Heterogeneity",
                     values = c(plot_colors)) +
  facet_wrap(~power,
             labeller = labeller(power = power.labs))+
  theme_bw()

@@

```{r corrected-figure, echo = F, fig.cap="Corrected projected sample sizes for variability and power levels. "}
decay <- 1-sqrt((summary_long$original_n-20)/summary_long$original_n)^log2(summary_long$original_n)
summary_long$new_sample <- summary_long$sample_size*decay
ggplot(summary_long %>% 
         group_by(original_n, power, variability) %>% 
         summarize(sample_size_avg = mean(sample_size),
                   new_sample_size_avg = mean(new_sample)),
       aes(original_n, new_sample_size_avg, color = variability)) + 
  geom_line(aes(original_n, original_n), color = "black") +
  geom_point(aes(original_n, original_n), color = "black") + 
  geom_point() + 
  geom_line() + 
  xlab("\nPilot Sample Size") + 
  ylab("Projected Sample Size\n") + 
  scale_color_manual(name = "Item Heterogeneity",
                     values = c(plot_colors)) +
  facet_wrap(~power,
             labeller = labeller(power = power.labs))+
  theme_bw()
```

@@

Last, we examined the question of an appropriate SE decile. First, the 0%, 10%, and 20% deciles are likely too restrictive, providing very large estimates that do not always find a reasonable sample size in proportion to the pilot sample size, scale size, and heterogeneity. If we examine the $R^2$ values for each decile of our regression equation separately, we find that the values are all $R^2$ \> .99 with very little differences between them. Figure \@ref(fig:decile40-figure) and \@ref(fig:decile50-figure) illustrates the corrected scores for simulations at the 40% and 50% decile recommended cutoff for item standard errors. For small heterogeneity, differences in decile are minimal, while larger heterogeneity shows more correction at the 40% decile range, especially for scales with larger potential variance. Therefore, we would suggest the 40% decile to overpower each item to determine the minimum and maximum sample size.

```{r decile40-figure, fig.cap = "Comparison of the cutoffs for 40% deciles across heterogeneity (columns), powering of items (rows), and scale size (color)."}
summary_long %>% 
  filter(name == "Decile 40") %>% 
  group_by(variability, original_n, scale_size, name, power) %>% 
  summarize(new_sample = mean(new_sample)) %>% 
  ggplot(., aes(original_n, new_sample, color = scale_size)) + 
  geom_point(aes(original_n, original_n), color = "black") + 
  geom_line(aes(original_n, original_n), color = "black") + 
  geom_point() + 
  geom_line() + 
  theme_classic(base_size = 10) + 
  facet_grid(power~ variability,
             labeller = labeller(power = power.labs)) +
  ggtitle("Results for 40% deciles") +
  xlab("\nPilot Sample Size") + 
  ylab("Corrected Sample Size\n") + 
  scale_color_manual(name = "Scale Size",
                     values = c(plot_colors)) + 
  theme_bw() +
  theme(legend.position="bottom")

```{r decile50-figure, fig.cap = "Comparison of the cutoffs for 50% deciles across heterogeneity (columns), powering of items (rows), and scale size (color)."}
summary_long %>% 
  filter(name == "Decile 50") %>% 
  group_by(variability, original_n, scale_size, name, power) %>% 
  summarize(new_sample = mean(new_sample)) %>% 
  ggplot(., aes(original_n, new_sample, color = scale_size)) + 
  geom_point(aes(original_n, original_n), color = "black") + 
  geom_line(aes(original_n, original_n), color = "black") + 
  geom_point() + 
  geom_line() +
  theme_classic(base_size = 10) + 
  facet_grid(power~ variability,
             labeller = labeller(power = power.labs)) + 
  ggtitle("Results for 40% deciles") +
  xlab("\nPilot Sample Size") + 
  ylab("Corrected Sample Size\n") + 
  scale_shape_discrete(name = "Scale Size") +
  scale_color_manual(name = "Cutoff Decile",
                     values = c(plot_colors)) + 
  theme_bw() +
  theme(legend.position="bottom")

@@

```{r sim_pop, include = T, echo = T}
devtools::install_github("SemanticPriming/semanticprimeR")
library(semanticprimeR)
pops <- simulate_population(mu = 4, # item means
  mu_sigma = .2, # variability in item means 
  sigma = 2, # item standard deviations
  sigma_sigma = .2, # standard deviation of the standard deviations
  number_items = 30, # number of items
  number_scores = 20, # number of participants
  smallest_sigma = .02, #* smallest possible standard deviation
  min_score = 1, #* minimum score for truncating purposes
  max_score = 7, #* maximum score for truncating purposes
  digits = 0) #* number of digits for rounding
